---
title: "Project Time series"
author: "Muchigi"
date: "8/4/2020"
output: html_document
---

1. Bring in the libraries

```{r }

library(tswge)
library(dplyr)
library(magrittr)
library(ggplot2)
library(readxl)
library(vars)
library(nnfor)
library(RColorBrewer)
```

# Prepare the data


```{r }
# filtering the data to United states because the data has all the continents
df1<-read.csv("C:\\Users\\pro3\\Google Drive\\Time series\\DATA\\owid-covid-data.csv")

df1<-filter(df1,location=='United States')


#changed the data type column to date
df1$Date <- as.Date(df1$date , format = "%m/%d/%y")


#gave  new case data the variable test positive
Test_positive = df1$new_cases[78:198]

#gave  new test data the variable Newtest
Newtest  = df1$new_tests [78:198]


#calculated positive percentage
positive_percent <- Test_positive/Newtest

# removed all the N/A's
positive_percent[is.na(positive_percent)] <- 0



```

## Visualized

plotted newcases , new deaths, and positive rate

```{r }
# created a line graph for new cases
ggplot(df1, aes(x=Date,y=new_cases))+geom_line(color="#69b3a2")+scale_x_date(limits = as.Date(c('2020-01-01','2020-07-20')))+ggtitle("New Cases Trend")

```

```{r }

#created a line graph for new deaths
ggplot(df1, aes(x=Date,y=new_deaths))+geom_line(color="#69b3a2")+scale_x_date(limits = as.Date(c('2020-01-01','2020-07-20')))+ggtitle("New Deaths Trend")


```


```{r }

#created a line graph for positive rate ratio
plot(positive_percent, type='l')

```




#Univariate Analysis

First will check the sample plots 

- we can see wondering behavior in the realisation
- there is slowly damping in the autocorrelation
- and we have a peak a zero in the spectral density
- this is an indication we have some non stationary factor


  


```{r }

#create sample plot for new cases 
plotts.sample.wge(Test_positive)


```


Next step was to overfit the models with an AR 12 and to check the factor table
```{r }

#Over fit with an Ar 12, will use the burg to fit the model 
est.ar.wge(Test_positive,p=12,type="burg")

```


Since we so evidence of a ARIMA  with 1 unit root  we decided to difference the data once

```{r }

#Over fit with an Ar 12, will use the burg to fit the model 
tp2_1=artrans.wge(Test_positive,phi.tr = 1)

```




Since the realizations look like we have some seasonlity, we over fitted the data again with an Ar 16 and  an Ar 12 to look at the factor tables.

We are look for similar models

```{r }

#Over fit with an Ar 12, will use the burg to fit the model 
est.ar.wge(tp2_1,p=12,type="burg")
```


```{r }

#Over fit with an  Ar16, will use the burg to fit the model 
est.ar.wge(tp2_1,p=16,type="burg")
```

looking at the factor tables we thought that they behaved like an AR 7

so we looked at the factor table for AR7
and we so saw some similarities
```{r }

#factor table for AR7
factor.wge(phi=c(rep(0,6),1))

```


After determining that its an AR7, we differenced the data
```{r }

#factor table for AR7
tp2_1_7=artrans.wge(tp2_1,phi.tr = c(rep(0,6),1))

```

Since the model looks stationary we used the AIC.5 to pick out model.
we are going to use AIC AICC and BIC

```{r }

#AIC
aic5.wge(tp2_1_7,p=0:10,q=0:3)

```

```{r }

#AICC
aic5.wge(tp2_1_7,p=0:10,q=0:3,type ="aicc")

```



```{r }

#BIC
aic5.wge(tp2_1_7,p=0:10,q=0:3,type ="bic")
```


Fit the model with the BIC model Ar(6)


```{r }

#fit the model with AR(6) 
est_tp2_1_7=est.arma.wge(tp2_1_7,p=6,q=0)

est_tp2_1_7$phi
```

Look for noise in our residual

 - Autocorrelation shows that the residuals are white noise
 - And the Ijung test failed to reject that the residuals are not white noise
 
    - this is because at K = 24 pval is 0.1879943
    - this is because at K = 48 pval is 0.8812295
```{r }

#code plots the sample plots
plotts.sample.wge(est_tp2_1_7$res,arlimits = T)

#ijung test for white noise
ljung.wge(est_tp2_1_7$res,p=6,q=0)
ljung.wge(est_tp2_1_7$res,p=6,q=0,K=48)

```

The code below calculates the ASE

```{r }

#variable for total length of the data
total=length(Test_positive)

#days length variable
length=15

#Here we forecast 15 days ahead and cross validation
f_tp2_1_7=fore.aruma.wge(Test_positive,phi=est_tp2_1_7$phi,theta = est_tp2_1_7$theta,s=7,d=1,n.ahead = length,lastn = T)


```

Forecast seven days ahead
```{r }
#below is the code for calculating the ASE
ASE.aruma=mean((f_tp2_1_7$f-Test_positive[(total-length+1):total])^2)
ASE.aruma
```



Forecast seven days ahead
```{r }
#forecast seven days ahead
Forecast_7days=fore.aruma.wge(df1$new_cases[78:198],phi=est_tp2_1_7$phi,theta = est_tp2_1_7$theta,s=7,d=1,n.ahead = 7,lastn = F)

```


Forecast 3 months ahead
```{r }
#forecast seven days ahead
Forecast_7days=fore.aruma.wge(df1$new_cases[78:198],phi=est_tp2_1_7$phi,theta = est_tp2_1_7$theta,s=7,d=1,n.ahead = 90,lastn = F)

```




#Univatiate MLP model

```{R}
#creates out train and test dataset
ml_newcase_train = ts(df1$new_cases[78:180])
ml_newcase_test = ts(df1$new_cases[181:198])
```

Below is where we fitted our model 

```{r}
fit3=mlp(ml_newcase_train,difforder = c(7),reps=50,comb="mean",allow.det.season = F)

fit3
```

After we fitted our model we ploted our actuals with the forecast data to see how well the model did

```{r}
fore.mlp =forecast(fit3 ,h=18)
plot(fore.mlp)

```


the plot below compares our forecast with our actual data  

```{r}
plot(df1$new_cases[181:198],type="l",)
points(seq(181:198),fore.mlp$mean,type='b',pch=15)

```

Calculated ASE for the mlp model  
```{r}
ASE.mlp = mean((df1$new_cases[181:198]-fore.mlp$mean)^2)
ASE.mlp
```

#Univariate Ensemble model

```{r}
r1=1-ASE.aruma/(ASE.mlp+ASE.aruma)
r2=1-ASE.mlp/(ASE.mlp+ASE.aruma)  
ensemble  = r1*f_tp2_1_7$f + r2*fore.mlp$mean

Date<-df1$date
plot(Test_positive[78:198] ~ Date[78:198], xaxt = "n", type = "l",xlab="Date", ylab = "Daily Test Positive", main = "20 day Mass Covid Forecast")

axis(1, Date[78:198], format(Date[78:198], "%b %d"), cex.axis = .7)
lines( ensemble  ~ Date[78:198], type = "l", col = "green")

ASE = mean((Test_positive[78:198] - ensemble)^2)
ASE
```


#Multivariate Time Series

The next step is we are going to use a Multivariate model to predict new cases

Will first us a Vector model

```{r }


# below we are preparing the data  for newcases, Newdeaths and Newtest
Newcase1 <- df1$new_cases[1:180]

Newcase1[is.na(Newcase1)] <- 0

Newdeaths <- df1$new_deaths[1:180]

Newdeaths[is.na(Newdeaths)] <- 0

Newtest1 <- df1$new_tests[1:180]

Newtest1[is.na(Newtest1)]<-0

#After creating the variable we combined them for our model
x <-cbind(Newcase1,Newdeaths,Newtest1)




```


Used Varselect to pick out our model
 - Model picked a p = 12
```{r pressure, echo=FALSE}

# here we are making varselect pick out our model
VARselect(x,lag.max = 12,type = "const",season = NULL,exogen = NULL)



```


Predictes 18 days ahead
```{r}

lsfit2=VAR(x,p=12,type = 'const')

pred = predict(lsfit2,n.ahead = 18)

```

the plot below compares our forecast with our actual data  

```{r pressure, echo=FALSE}

plot(df1$new_cases[181:198],type="l",)
points(seq(181:198),pred$fcst$Newcase1[1:18,1],type='b',pch=15)

```

Got the ASE for the model

- it performed better than the univariate model
```{r pressure, echo=FALSE}
#ASE calculation
ASE=mean((pred$fcst$Newcase1[1:18,1]-df1$new_cases[181:198])^2)
ASE
```



Forecast seven days ahead
```{r }
pred = predict(lsfit2,n.ahead = 7)
fanchart(pred, colors =brewer.pal(n=8,name = "Blues"))

```


Forecast 90 days ahead
```{r }
pred = predict(lsfit2,n.ahead = 90)
fanchart(pred, colors =brewer.pal(n=8,name = "Blues"))

```



# MLP model

The below process is for Modeling the newcase dataset with ML to see if will get better results

Prepared the data

```{r }

#creates out train and test dataset
ml_newcase_train = ts(df1$new_cases[1:180])
ml_newcase_test = ts(df1$new_cases[181:198])

# here we combine out explainable variables
Combine=data.frame(Newdeaths1=ts(df1$new_deaths),Tests =ts(df1$new_tests))

#removed Na's from out test dataset
Combine$Tests[is.na(Combine$Tests)]<-0


```


Below is where we fitted our model 

```{r }

fit3=mlp(ml_newcase_train,difforder = c(7), xreg=Combine,reps=100,comb="mean")

fit3

```


After we fitted our model we ploted our actuals with the forecast data to see how well the model did
```{r }

fore.mlp =forecast(fit3,xreg =Combine ,h=18)
plot(fore.mlp)

```



plot actuals verses predicted realization
```{r}

plot(df1$new_cases[181:198],type="l",)
points(seq(181:198),fore.mlp$mean,type='b',pch=15)

```



Calculated the ASE for our mlp model

- the mlp perfored worse


```{r }

ASE = mean((df1$new_cases[181:198]-fore.mlp$mean)^2)
ASE
```






